An investigation into using scagnostics to find interesting projections of high-dimensional data

### 2018-10-23 scagnostics.Rmd
Cross referencing in Sec 3.

  - smoothness: the index function should evolve smoothly between nearby projections, i.e. on an interpolated path we expect to observe slow and consistent change in the index function. On the other hand if two similar projections produce very different index values, i.e. we observe sharp jumps when looking at the evolution of the index function, and in particular if those fluctuations dominate the overall evolution, the index is not suitable for projection pursuit.
  - speed: the computation time is an important factor in particular when the guided tour is used to view the optimisation in real time. Note that longer computation times may be acceptable when first recording the optimisation and replaying the guided tour once the maximum has been identified.
  - ability to find fine and broad structure: We consider a broad structure one that is roughly observable over a large angle around the optimal projection, it should in general be easy to identify and a good ppi should show smooth increase over a large number of projections as we are approaching the optimal view. On the other hand fine structures are difficult to identify as they are only visible close to the optimal view. This poses problems in particular for the optimization, but we may use index definitions to amplify larger index values while suppressing smaller ones, such that the optimization will be less sensitive to noise. \textcolor{red}{there are two different considerations, one is that described above, it relates to the concept of squint angle discussed in \cite{posse95b}. On the other hand should also consider detectin of "fine" structure vs noise, this would be controlled e.g. by the allowed number of bins for the MINE index functions, additionally this will impact computing times, i.e. trade accuracy for efficiency}
  - distribution of index values: the distribution of index values when studying data points without structure should have a low variance, and be clearly separated from index values obtained when looking at structured data distributions.
  - rotation invariance: in the tour there is an assumed degeneracy between all projections obtained when rotating within the projected space only, for 2-d projections it means that we cannot assing x- and y- axis without ambiguity. A projection pursuit index should therefore be invariant under rotation of the basis in the projection plane. By costruction this is not the case for most index functions defined above. (One way around could be profiling the function over all possible rotations in the plane, howevere computing time may be a major concern in this case.) (Other examples where rotation invariance is important include image recognition and texture classification, rotation angle may be built in parameter for summary calculation, e.g. https://doi.org/10.1016/S0146-664X(75)80008-6, other example is interpolating over angles when building NN, see e.g. https://arxiv.org/pdf/1805.12301.pdf)

@@

## Smoothness
We first study the smoothness of each measure calculated on a sequence of 2-d projections obtained via an interpolated planned tour path moving between the projection onto V1-V2 to the projection onto V3-V4 and using the default interpolation angle of 0.05. Each time step refers to an interpolation step. Since the value of the TIC index is strongly dependent on the input distribution sample size we have normalised the TIC measure by its maximum for each individual dataset, all other measures are reproduced as reported by the calculation. The results are shown in Figure \ref{fig:plotV1V2toV3V4}. The standard index functions "holes" and "cmass" show the desired behaviour in that the change is smooth over interpolated views ((it is however curious to see the actual evolution, which seems not to depend on the underlying random distributions, but is clearely evolving over the path of the planned tour)). Considering now the indices from the scagnostics family, we first note stronge dependence on the sample size. The convex index is largely sensitive to random fluctuations when considering small samples (p=100), while being mostly flat for larger samples, especially in the case of an underlying uniform distribution. The skinny index on the other hand is systematically larger for the smaller samples, while stringy appears not to be sensitive to the sample size, but resulting in large fluctuations and sharp jumps for all datasets. Distance correlation and the spline based index show similar behaviour, i.e. when considering large enough data samples (p=1000) the values are consistently close to zero, as expected for a distribution without structure. On the other hand somewhat larger values (but still far from the maximum value of 1) are found when considering small datasets, where distributions are however mostly smooth, with few kinks depending on index function and underlying distribution. Finally the information based metrics also show systematic differences based on the sample size, with sampling fluctuations being significant for small datasets. Note here that fluctuations present in Figure \ref{fig:plotV1V2toV3V4} may be amplified, as maximum values for most index functions are far from the theoretic maximum.

@@

## Distributions of index values

Compare typical index values over the two tour paths. Notation: path 1 / 2 labels tour from V1-V2 to V3-V4 / V5-V6. Typically find broader distributions for small samples because of the large fluctuations in this case. Some indices show preferred behaviour, i.e. sharp peaks around low values for path 1, and very broad distributions leading up to maximum values for path 2. For other indices we can see the strong dependence of the index on sample size, the most extreme example is the skinny index.

@@

## Rotation dependence
\textcolor{red}{testin rotation dependence by turning sine dataset with $\alpha \in [0,2\pi]$}
Interesting to note that some indexes are symmetric around $\pi$, but scagnostics indexes are very arbitrary, dependence is introduced via binnibng (and outlier removal?) here. Observe huge drops for some of the indexes! 
```{r rotationDep, fig.width=6, fig.height=7, fig.cap="Rotation dependence of the considered index functions tested on the sine 1000 dataset, by rotating the special view by an angle alpha."}
sineM <- as.matrix(select(sin1000, V5, V6))
sc <- tibble(
    holes=numeric(),
    cmass=numeric(),
    convex=numeric(),
    skinny=numeric(),
    stringy=numeric(),
    dcor2d=numeric(),
    splines2d=numeric(),
    MIC=numeric(),
    TIC=numeric(),
    alpha=numeric()
)
for (a in seq(0,2*pi, pi/20)){
  rotM <- matrix(c(cos(a), sin(a), -sin(a), cos(a)), ncol = 2)
  dprj <- sineM %*% rotM
  scagRes <- scagnostics(dprj)
  dcorRes <- dcor2d(dprj[,1], dprj[,2])
  splineRes <- splines2d(dprj[,1], dprj[,2])
  mineRes <- mine(dprj[,1], dprj[,2])
  holesRes <- holes(dprj)
  cmassRes <- cmass(dprj)
  sc <- add_row(sc, holes=holesRes, cmass=cmassRes,
                convex=scagRes[,"Convex"], skinny=scagRes[,"Skinny"],
                stringy=scagRes[,"Stringy"], dcor2d=dcorRes,
                splines2d=splineRes, MIC=mineRes$MIC,
                TIC=mineRes$TIC, alpha=a)
  }

@@

## Note on calculation time
Computing time differs for each index, and depends strongly on the number of data points (as well as probably some dependence on distribution type), here we use as an example the 6-d sphere and calculate the computing time for different indices over 100 interpolated grand tour projections as a function of the sample size.
We see that the evaluation of the spline index shows weak dependence on the sample size, making it the least efficient for very small data sets, but one of the fastest when considering large data sets. The situation is similar for calculating the scagnositics measures, which are however faster to compute on datasets of around 10 points. This is because the implementation is first binning the data points, which is fast compared to the evaluation that requires triangulation of the binned points. The distance correlation based measure exhibits strong increase with sample size, we note however that the implementation also allowes for evaluation on binned data points which we have not considered. Finally for the MINE family the computing time increases exponentially as can be expected from the dependence on the sample size, where we see that the improved $MIC_e$ algorithm is somewhat faster than the original implementation, see also @JMLRv1715308.

@@

>>>>

  - smoothness: the index function should evolve smoothly between nearby projections, i.e. on an interpolated path we expect to observe slow and consistent change in the index function. On the other hand if two similar projections produce very different index values, i.e. we observe sharp jumps when looking at the evolution of the index function, and in particular if those fluctuations dominate the overall evolution, the index is not suitable for projection pursuit. We evaluate smoothness both for transitions between noise variables and when moving from noise variables to an informative view in Section \ref{sec:smooth}.
  - speed: the computation time is an important factor in particular when the guided tour is used to view the optimisation in real time. Note that longer computation times may be acceptable when first recording the optimisation and replaying the guided tour once the maximum has been identified. Computation times for some benchmark scenarios are compared in Section \ref{sec:timer}.
  - ability to find fine and broad structure: We consider a broad structure one that is roughly observable over a large angle around the optimal projection, it should in general be easy to identify and a good ppi should show smooth increase over a large number of projections as we are approaching the optimal view. On the other hand fine structures are difficult to identify as they are only visible close to the optimal view. This poses problems in particular for the optimization, but we may use index definitions to amplify larger index values while suppressing smaller ones, such that the optimization will be less sensitive to noise. \textcolor{red}{there are two different considerations, one is that described above, it relates to the concept of squint angle discussed in \cite{posse95b}. On the other hand should also consider detectin of "fine" structure vs noise, this would be controlled e.g. by the allowed number of bins for the MINE index functions, additionally this will impact computing times, i.e. trade accuracy for efficiency}. We address these questions in Section \ref{sec:params}
  - distribution of index values: the distribution of index values when studying data points without structure should have a low variance, and be clearly separated from index values obtained when looking at structured data distributions. \textcolor{red}{I dont really understand this part.} We show examples in Section \ref{sec:dist}
  - rotation invariance: in the tour there is an assumed degeneracy between all projections obtained when rotating within the projected space only, for 2-d projections it means that we cannot assing x- and y- axis without ambiguity. A projection pursuit index should therefore be invariant under rotation of the basis in the projection plane. By costruction this is not the case for most index functions defined above. \textcolor{red}{One way around could be profiling the function over all possible rotations in the plane, howevere computing time may be a major concern in this case. Other examples where rotation invariance is important include image recognition and texture classification, rotation angle may be built in parameter for summary calculation, e.g. https://doi.org/10.1016/S0146-664X(75)80008-6, other example is interpolating over angles when building NN, see e.g. https://arxiv.org/pdf/1805.12301.pdf)}. We illustrate the impact of non invariance in Section \ref{sec:rot}

@@

## Smoothness
\label{sec:smooth}
We first study the smoothness of each measure calculated on a sequence of 2-d projections obtained via an interpolated planned tour path moving between the projection onto V1-V2 to the projection onto V3-V4 and using the default interpolation angle of 0.05. Each time step refers to an interpolation step. Since the value of the TIC index is strongly dependent on the input distribution sample size we have normalised the TIC measure by its maximum for each individual dataset, all other measures are reproduced as reported by the calculation. The results are shown in Figure \ref{fig:plotV1V2toV3V4}. The standard index functions "holes" and "cmass" show the desired behaviour in that the change is smooth over interpolated views ((it is however curious to see the actual evolution, which seems not to depend on the underlying random distributions, but is clearely evolving over the path of the planned tour)). Considering now the indices from the scagnostics family, we first note stronge dependence on the sample size. The convex index is largely sensitive to random fluctuations when considering small samples (p=100), while being mostly flat for larger samples, especially in the case of an underlying uniform distribution. The skinny index on the other hand is systematically larger for the smaller samples, while stringy appears not to be sensitive to the sample size, but resulting in large fluctuations and sharp jumps for all datasets. Distance correlation and the spline based index show similar behaviour, i.e. when considering large enough data samples (p=1000) the values are consistently close to zero, as expected for a distribution without structure. On the other hand somewhat larger values (but still far from the maximum value of 1) are found when considering small datasets, where distributions are however mostly smooth, with few kinks depending on index function and underlying distribution. Finally the information based metrics also show systematic differences based on the sample size, with sampling fluctuations being significant for small datasets. Note here that fluctuations present in Figure \ref{fig:plotV1V2toV3V4} may be amplified, as maximum values for most index functions are far from the theoretic maximum.

@@

## Distributions of index values
\label{sec:dist}
Compare typical index values over the two tour paths. Notation: path 1 / 2 labels tour from V1-V2 to V3-V4 / V5-V6. Typically find broader distributions for small samples because of the large fluctuations in this case. Some indices show preferred behaviour, i.e. sharp peaks around low values for path 1, and very broad distributions leading up to maximum values for path 2. For other indices we can see the strong dependence of the index on sample size, the most extreme example is the skinny index.

@@

## Rotation dependence
\label{sec:rot}
\textcolor{red}{testin rotation dependence by turning sine dataset with $\alpha \in [0,2\pi]$}
Interesting to note that some indexes are symmetric around $\pi$, but scagnostics indexes are very arbitrary, dependence is introduced via binnibng (and outlier removal?) here. Observe huge drops for some of the indexes! 
```{r rotationDep, fig.width=6, fig.height=7, fig.cap="Rotation dependence of the considered index functions tested on the sine 1000 dataset, by rotating the special view by an angle alpha."}
sineM <- as.matrix(select(sin1000, V5, V6))
sc <- tibble(
    holes=numeric(),
    cmass=numeric(),
    convex=numeric(),
    skinny=numeric(),
    stringy=numeric(),
    dcor2d=numeric(),
    splines2d=numeric(),
    MIC=numeric(),
    TIC=numeric(),
    alpha=numeric()
)
for (a in seq(0,2*pi, pi/20)){
  rotM <- matrix(c(cos(a), sin(a), -sin(a), cos(a)), ncol = 2)
  dprj <- sineM %*% rotM
  scagRes <- scagnostics(dprj)
  dcorRes <- dcor2d(dprj[,1], dprj[,2])
  splineRes <- splines2d(dprj[,1], dprj[,2])
  mineRes <- mine(dprj[,1], dprj[,2])
  holesRes <- holes(dprj)
  cmassRes <- cmass(dprj)
  sc <- add_row(sc, holes=holesRes, cmass=cmassRes,
                convex=scagRes[,"Convex"], skinny=scagRes[,"Skinny"],
                stringy=scagRes[,"Stringy"], dcor2d=dcorRes,
                splines2d=splineRes, MIC=mineRes$MIC,
                TIC=mineRes$TIC, alpha=a)
  }

@@

## Note on calculation time
\label{sec:timer}
Computing time differs for each index, and depends strongly on the number of data points (as well as probably some dependence on distribution type), here we use as an example the 6-d sphere and calculate the computing time for different indices over 100 interpolated grand tour projections as a function of the sample size.
We see that the evaluation of the spline index shows weak dependence on the sample size, making it the least efficient for very small data sets, but one of the fastest when considering large data sets. The situation is similar for calculating the scagnositics measures, which are however faster to compute on datasets of around 10 points. This is because the implementation is first binning the data points, which is fast compared to the evaluation that requires triangulation of the binned points. The distance correlation based measure exhibits strong increase with sample size, we note however that the implementation also allowes for evaluation on binned data points which we have not considered. Finally for the MINE family the computing time increases exponentially as can be expected from the dependence on the sample size, where we see that the improved $MIC_e$ algorithm is somewhat faster than the original implementation, see also @JMLRv1715308.

@@

## Index function parameters
\label{sec:params}
Some parameters must be fixed when using the index functions introduced in ...
